---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

Show any code that is needed to load the data and process/transform it into a
format suitable for analysis.

```{r load_data, echo = TRUE}
# Unzip the data file if it hasn't been unzipped yet, then read it in
if (!file.exists("activity.csv") && file.exists("activity.zip")) {
    unzip("activity.zip")
}

activity <- read.csv("activity.csv", colClasses = c("integer", "character", "integer"))

# Convert the date column to Date class
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")

str(activity)
```

## What is mean total number of steps taken per day?

For this part of the assignment, the missing values in the dataset are ignored.

### Total number of steps taken per day

```{r total_steps, echo = TRUE}
total_steps_per_day <- aggregate(steps ~ date, data = activity, sum, na.rm = TRUE)
head(total_steps_per_day)
```

### Histogram of the total number of steps taken each day

```{r histogram_total_steps, echo = TRUE}
hist(total_steps_per_day$steps,
     main = "Histogram of Total Number of Steps Taken Each Day",
     xlab = "Total steps per day",
     col = "steelblue",
     breaks = 20)
```

### Mean and median number of steps taken per day

```{r mean_median_steps, echo = TRUE}
mean_steps   <- mean(total_steps_per_day$steps)
median_steps <- median(total_steps_per_day$steps)

mean_steps
median_steps
```

The mean total number of steps taken per day is **`r format(mean_steps, big.mark = ",", nsmall = 2)`**
and the median is **`r format(median_steps, big.mark = ",")`**.

## What is the average daily activity pattern?

### Time series plot of the average number of steps taken

```{r time_series_plot, echo = TRUE}
interval_avg <- aggregate(steps ~ interval, data = activity, mean, na.rm = TRUE)

plot(interval_avg$interval, interval_avg$steps, type = "l",
     main = "Average Number of Steps Taken Across All Days",
     xlab = "5-minute interval",
     ylab = "Average number of steps",
     col = "steelblue", lwd = 2)
```

### The 5-minute interval with the maximum average number of steps

```{r max_interval, echo = TRUE}
max_interval <- interval_avg[which.max(interval_avg$steps), ]
max_interval
```

On average across all the days in the dataset, the 5-minute interval
**`r max_interval$interval`** contains the maximum number of steps
(`r round(max_interval$steps, 2)` steps on average).

## Imputing missing values

### Total number of missing values

```{r count_na, echo = TRUE}
total_na <- sum(is.na(activity$steps))
total_na
```

There are **`r total_na`** rows with missing (`NA`) step values in the dataset.

### Strategy for filling in missing values

The strategy used here is to fill in each missing value with the **mean number
of steps for that 5-minute interval**, computed across all days (the same
`interval_avg` table calculated above).

### Create a new dataset with the missing data filled in

```{r impute_data, echo = TRUE}
activity_imputed <- activity

# Match each row's interval to the corresponding average and substitute it in
na_rows <- is.na(activity_imputed$steps)
activity_imputed$steps[na_rows] <- interval_avg$steps[match(activity_imputed$interval[na_rows],
                                                              interval_avg$interval)]

# Confirm there are no remaining NAs
sum(is.na(activity_imputed$steps))
```

### Histogram of total steps per day (imputed data), with mean and median

```{r histogram_imputed, echo = TRUE}
total_steps_imputed <- aggregate(steps ~ date, data = activity_imputed, sum)

hist(total_steps_imputed$steps,
     main = "Histogram of Total Steps Per Day (Missing Values Imputed)",
     xlab = "Total steps per day",
     col = "darkorange",
     breaks = 20)

mean_steps_imputed   <- mean(total_steps_imputed$steps)
median_steps_imputed <- median(total_steps_imputed$steps)

mean_steps_imputed
median_steps_imputed
```

**Do these values differ from the estimates in the first part of the assignment?**

```{r compare_estimates, echo = TRUE}
data.frame(
    metric   = c("mean", "median"),
    original = c(mean_steps, median_steps),
    imputed  = c(mean_steps_imputed, median_steps_imputed)
)
```

The mean stays essentially the same (`r round(mean_steps, 2)` vs.
`r round(mean_steps_imputed, 2)`), because entire missing days were imputed
using each interval's average, which reproduces a "typical" day and does not
pull the overall daily mean away from the original estimate. The median
shifts very slightly upward, from `r median_steps` to
`r round(median_steps_imputed, 2)`, because one of the imputed days (built
entirely from interval averages) now sits at the exact mean value and becomes
the middle observation. Overall, the impact of imputing missing data on the
estimates of total daily steps is small; it mainly reduces the height of the
lowest bar in the histogram (the "zero-step" days that were previously just
missing) since those days now show a plausible average day of steps rather
than being dropped from the daily totals.

## Are there differences in activity patterns between weekdays and weekends?

### Create a factor variable indicating weekday or weekend

```{r weekday_weekend, echo = TRUE}
activity_imputed$day_type <- factor(
    ifelse(weekdays(activity_imputed$date) %in% c("Saturday", "Sunday"),
           "weekend", "weekday")
)

table(activity_imputed$day_type)
```

### Panel plot: average steps per interval, weekday vs. weekend

```{r panel_plot, echo = TRUE}
interval_daytype_avg <- aggregate(steps ~ interval + day_type,
                                   data = activity_imputed, mean)

par(mfrow = c(2, 1), mar = c(4, 4, 3, 2))

with(subset(interval_daytype_avg, day_type == "weekday"),
     plot(interval, steps, type = "l", col = "steelblue", lwd = 2,
          main = "Weekday", xlab = "Interval", ylab = "Avg. steps"))

with(subset(interval_daytype_avg, day_type == "weekend"),
     plot(interval, steps, type = "l", col = "darkorange", lwd = 2,
          main = "Weekend", xlab = "Interval", ylab = "Avg. steps"))
```

Activity on weekdays shows a sharp peak in the morning (around interval
835, roughly 8:35 am), consistent with a morning commute or exercise
routine, followed by lower and fairly flat activity through the rest of the
day. Weekend activity is more spread out and starts a bit later in the day,
with a lower single morning peak but sustained higher activity levels
throughout the late morning and afternoon compared to weekdays.
